---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(car)
library(MASS)
```

# Notes

- For dental insurance, should "Don't know" be counted as 0 or 1 in binary version?

- brush_time option 1 being "Unsure" likely is giving issues

- All models, unless they are filtered for them already, control for gender and race. Any others?

- Some models are logistic regression models, due to a binary outcome variable

- Some variables are considered numeric but could be counted as ordinal variables
  - dental_visits
    - Just has a 3+ category, so likely not worth the extra complexity
  - mouth_pain
    - Categories are not evenly spaced, so ordinal logistic regression may make sense
    - However, if the category spacing makes sense, we could keep a linear regression model
    - The data is right skewed, but that doesn't necessarily mean the category spacing is bad
    - The ordinal logistic regression model has good (but not great) 95 percent CIs
  - brushing
    - Could work either way, probably better as ordinal
  - brush_time
    - Could work either way, as the gaps are relatively even (except for level 1)
  - Overall, ordinal logistic regression, compared to linear regression, has a more complex interpretation

- Race variable has a couple groups with 1 patient
  - Recommend grouping into the "Other" group

- For race/ethnicity among women (number 2 in the document), difficult to make groups
  - Some women belong in multiple of the groups

- Regarding the Oral Health Night section (number 3 in the document): 
  - Can't find a variable for attending Oral Health Night
  - referral_date seems like it might be correct, but that is not in the data dictionary

- For the section about oral hygiene habits and dental care access:
  - Another variable is needed to complete a t-test or regression model

```{r}
# Load data
oh <- read_csv("C:/Users/hebers/OneDrive - University Of Massachusetts Medical School/Documents/UMass Projects/OH-QI-Pajak/WFCCOralHealthSurvey_DATA_2024-07-18_1356_SH.csv")
oh
```

```{r}
# Transform data
# Already removed record ID 15, as it was blank
oh2 <- oh %>% 
  mutate(gendernew = NULL) %>% 
  mutate(racenew = NULL) %>% 
  mutate(dent_int_bin = NULL) %>% 
  mutate(englishnew = NULL) %>% 
  mutate(mouth_pain_fac = NULL)
oh2$dental_insurance <- factor(oh2$dental_insurance, labels = c("MassHealth", "Private", "Other", "Don't know", "No"))
for(i in 1:nrow(oh2)){
  oh2$gendernew[i] = ifelse(oh2$gender___1[i] == 1, "Female", ifelse(oh2$gender___2[i] == 1, "Male", ifelse(oh2$gender___3[i] == 1, "Trans female", ifelse(oh2$gender___4[i] == 1, "Trans male", ifelse(oh2$gender___5[i] == 1, "Non-binary", ifelse(oh2$gender___6[i] == 1, "Gender fluid", ifelse(oh2$gender___7[i] == 1, "Prefer not to answer", ifelse(oh2$gender___8[i] == 1, "Other", NA))))))))
}
for(i in 1:nrow(oh2)){
  oh2$racenew[i] = ifelse(oh2$race___1[i] == 1 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, "American Indian or Alaska Native", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 1 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, "Asian", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 1 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, "Black or African American", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 1 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, "Native Hawaiian or Other Pacific Islander", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 1 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, "White", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 1 & oh2$race___7[i] == 0, "Prefer not to answer", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 1, "Other", ifelse(oh2$race___1[i] == 0 & oh2$race___2[i] == 0 & oh2$race___3[i] == 0 & oh2$race___4[i] == 0 & oh2$race___5[i] == 0 & oh2$race___6[i] == 0 & oh2$race___7[i] == 0, NA, "Multiracial"))))))))
}
oh2$dent_int_bin <- ifelse(oh2$dental_insurance == "No", 0, 1)
oh2$englishnew <- ifelse(oh2$language == 4, 0, 1)
oh2$mouth_pain_fac <- ordered(as.factor(oh2$mouth_pain))
oh2
```

# Models comparing yes/no dental insurance

```{r}
# Model outcome: Mouth pain in last 12 months
m1 <- lm(mouth_pain ~ dent_int_bin + gendernew + racenew, data = oh2)
summary(m1)
```

```{r}
# m1, but with mouth_pain ordered factor version
m1fac <- polr(mouth_pain_fac ~ dent_int_bin + gendernew + racenew, data = oh2, Hess = TRUE)
summary(m1fac)
```

```{r}
# m1fac ORs and CIs
exp(cbind(OR = coef(m1fac), confint(m1fac)))
```

```{r}
# Outcome: Oral health care received in last 12 months
m2 <- glm(dental_visits ~ dent_int_bin + gendernew + racenew, data = oh2)
summary(m2)
```
```{r}
# Outcome: Oral health care received in last 12 months, Poisson distribution
# Poisson distribution compared, due to outcome being count data
m2.poisson <- glm(dental_visits ~ dent_int_bin + gendernew + racenew, family = "poisson", data = oh2)
summary(m2.poisson)
```

```{r}
# Outcome: Needing to access dental care but unable to
m3 <- glm(access_dental_care ~ dent_int_bin + gendernew + racenew, family = "binomial", data = oh2)
summary(m3)
```

```{r}
# Outcome: Frequency of teeth brushing
m4 <- lm(brushing ~ dent_int_bin + gendernew + racenew, data = oh2)
summary(m4)
```

```{r}
# Outcome: Duration of teeth brushing
m5 <- lm(brush_time ~ dent_int_bin + gendernew + racenew, data = oh2)
summary(m5)
```

```{r}
# Outcome: Dental floss usage
m6 <- lm(flossing ~ dent_int_bin + gendernew + racenew, data = oh2)
summary(m6)
```

```{r}
# Outcome: Need for dental hygiene supplies to bring home
m7 <- glm(dental_supplies ~ dent_int_bin + gendernew + racenew, family = "binomial", data = oh2)
summary(m7)
```

# Women, dental insurance, and race/ethnicity

```{r}
# Some overlap for race and ethnicity
ohwomen <- oh2 %>% 
  filter(gendernew == "Female")
table(ohwomen$racenew, ohwomen$ethnicity)
# It seems there are 3 black/African American women and 1 white woman who also
#   identify as Hispanic/Latino
```

# Women, dental insurance, and English proficiency

```{r}
# Keep in mind that englishnew has 0 as no English, 1 as some or more English
oh.hispanic.women <- oh2 %>% 
  filter(gendernew == "Female" & ethnicity == 1)
oh.black.women <- oh2 %>% 
  filter(gendernew == "Female" & racenew == "Black or African American")
```

```{r}
# Outcome: English proficiency
# Hispanic/Latino women
m8 <- glm(englishnew ~ dent_int_bin, family = "binomial", data = oh.hispanic.women)
summary(m8)
```

```{r}
# Outcome: English proficiency
# Black or African American women
m9 <- glm(englishnew ~ dent_int_bin, family = "binomial", data = oh.black.women)
summary(m9)
```

# Women, Oral Health Night, and dental insurance

```{r}
# See notes
# Could not find Oral Health Night variable
```

# Women who brush often and use floss and access to care

```{r}
# Filtering for: 
#  - Women
#  - Hispanic ethnicity
#  - Brushing at least daily
#  - Brushing for at least one minute
#  - Flossing at occasionally

# Only 9 people out of 84 fit in this group
oh.hispanic.women.brush.floss <- oh2 %>% 
  filter(gendernew == "Female" & 
           ethnicity == 1 & 
           brushing >= 5 & 
           brush_time >= 4 & 
           flossing >= 2)
```

```{r}
# Breakdown of access_dental_care variable
table(oh.hispanic.women.brush.floss$access_dental_care)
```

